---
title: "P1 Add Health Depression Measurement Comparison Analysis Notebook"
author: "Rophence Ojiambo"
date: "`r Sys.Date()`"
output:
  html_notebook: default
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---


# Overview

This notebook combines the full P1 analysis workflow into one reproducible R Notebook.

Execution order:

1. Construct the three depressive symptom measures.
2. Build the locked analytic sample.
3. Produce descriptive statistics, reliability estimates, Figure 1, and 80th-percentile classification thresholds.
4. Produce subgroup comparisons and Figure 3.
5. Run regression models and Figure 4.
6. Compute classification agreement and Figure 2.

This notebook assumes the existing project folders are already present:

```{r folder-check, eval=FALSE}
list.files()
file.exists("data")
file.exists("outputs")
file.exists("outputs/figures")
```

The notebook does not create new folders. Data are read from `data/`, saved datasets are written to `data/`, and output tables/figures are written to `outputs/` and `outputs/figures/`.

# Setup

```{r setup, message=FALSE, warning=FALSE}
rm(list = ls())

library(tidyverse)
library(writexl)
library(psych)
library(broom)
library(ggplot2)


################################################################################
# Shared figure theme
################################################################################

project_figure_theme <- theme_minimal(
  base_size = 14,
  base_family = "sans"
) +
  theme(
    # All theme-controlled text
    text = element_text(
      face = "bold",
      colour = "black"
    ),

    # Titles and subtitles
    plot.title = element_text(
      hjust = 0.5,
      size = 18,
      face = "bold",
      colour = "black",
      margin = margin(b = 8)
    ),
    plot.subtitle = element_text(
      hjust = 0.5,
      size = 13,
      face = "bold",
      colour = "black",
      margin = margin(b = 8)
    ),

    # Axis titles
    axis.title = element_text(
      size = 14,
      face = "bold",
      colour = "black"
    ),
    axis.title.x = element_text(
      margin = margin(t = 8)
    ),
    axis.title.y = element_text(
      margin = margin(r = 8)
    ),

    # Axis tick labels
    axis.text = element_text(
      size = 11,
      face = "bold",
      colour = "black"
    ),
    axis.text.x = element_text(
      hjust = 0.5
    ),
    axis.text.y = element_text(
      margin = margin(r = 6)
    ),

    # Facet headings
    strip.text = element_text(
      size = 12,
      face = "bold",
      colour = "black"
    ),

    # Legends
    legend.position = "bottom",
    legend.title = element_text(
      size = 12,
      face = "bold",
      colour = "black"
    ),
    legend.text = element_text(
      size = 11,
      face = "bold",
      colour = "black"
    ),

    # Captions
    plot.caption = element_text(
      size = 9,
      face = "bold",
      colour = "black",
      hjust = 1,
      margin = margin(t = 8)
    ),

    # Grid and margins
    panel.grid.minor = element_blank(),
    plot.margin = margin(
      t = 15,
      r = 20,
      b = 15,
      l = 20
    )
  )

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```


# Step 1: Construct depressive symptom measures

```{r step-01}
################################################################################
# Script: 01_construct_measures_Rophence.R
# Purpose: Construct depressive symptom measures for the Add Health P1 project
################################################################################



# Extract the actual Add Health numeric code from factor-labelled variables.
# Example: "(95) (95) 1995" becomes 95.
get_code <- function(x) {
  readr::parse_number(as.character(x))
}

# Load the Add Health Wave I data
loaded_name <- load("data/21600-0001-Data.rda")

# Save the loaded dataset with a clear working name
addhealth_raw <- get(loaded_name[1])


################################################################################
# 1. Define depressive symptom items
################################################################################

# Full modified CES-D items
cesd_items <- paste0("H1FS", 1:19)

# Positive affect items that must be reverse-coded
positive_affect_items <- c("H1FS4", "H1FS8", "H1FS11", "H1FS15")

# Four items used for the brief screener
brief_items <- c("H1FS6", "H1FS16", "H1FS3", "H1FS13")


################################################################################
# 2. Clean depressive symptom items
################################################################################

# Valid H1FS responses are 0, 1, 2, and 3.
# Special codes 6 = refused and 8 = don't know are set to missing.
clean_h1fs <- function(x) {
  x_num <- get_code(x)
  
  case_when(
    x_num %in% 0:3     ~ x_num,
    x_num %in% c(6, 8) ~ NA_real_,
    TRUE               ~ NA_real_
  )
}

# Clean all 19 depressive symptom items
addhealth_measures <- addhealth_raw %>%
  mutate(
    across(
      all_of(cesd_items),
      clean_h1fs,
      .names = "{.col}_clean"
    )
  )


################################################################################
# 3. Reverse-code positive affect items
################################################################################

# Original scale: 0 to 3
# Reverse-coded scale: 3 - original value
# Higher values should always mean more depressive symptoms.
addhealth_measures <- addhealth_measures %>%
  mutate(
    H1FS4_rev  = 3 - H1FS4_clean,
    H1FS8_rev  = 3 - H1FS8_clean,
    H1FS11_rev = 3 - H1FS11_clean,
    H1FS15_rev = 3 - H1FS15_clean
  )


################################################################################
# 4. Construct the three depressive symptom measures
################################################################################

# Items used in the full 19-item modified CES-D score
# The positive affect items use their reverse-coded versions.
full_score_items <- c(
  "H1FS1_clean",
  "H1FS2_clean",
  "H1FS3_clean",
  "H1FS4_rev",
  "H1FS5_clean",
  "H1FS6_clean",
  "H1FS7_clean",
  "H1FS8_rev",
  "H1FS9_clean",
  "H1FS10_clean",
  "H1FS11_rev",
  "H1FS12_clean",
  "H1FS13_clean",
  "H1FS14_clean",
  "H1FS15_rev",
  "H1FS16_clean",
  "H1FS17_clean",
  "H1FS18_clean",
  "H1FS19_clean"
)

# Items used in the four-item brief screener
brief_score_items <- paste0(brief_items, "_clean")

# Pull item data for scoring
full_item_data <- addhealth_measures %>%
  select(all_of(full_score_items))

brief_item_data <- addhealth_measures %>%
  select(all_of(brief_score_items))

# Construct the three outcomes
addhealth_measures <- addhealth_measures %>%
  mutate(
    # Full scale is calculated only when all 19 items are non-missing.
    cesd_full_score = if_else(
      complete.cases(full_item_data),
      rowSums(full_item_data),
      NA_real_
    ),
    
    # Brief screener is calculated only when all 4 items are non-missing.
    cesd_brief_score = if_else(
      complete.cases(brief_item_data),
      rowSums(brief_item_data),
      NA_real_
    ),
    
    # Single-item depressed mood outcome.
    cesd_single_depressed_ord = H1FS6_clean
  )


################################################################################
# 5. Documentation and quality checks
################################################################################

# Documentation table
measure_documentation <- tibble(
  variable_name = c(
    "cesd_full_score",
    "cesd_brief_score",
    "cesd_single_depressed_ord"
  ),
  source_items = c(
    "H1FS1-H1FS19, with H1FS4, H1FS8, H1FS11, and H1FS15 reverse-coded",
    "H1FS6, H1FS16, H1FS3, H1FS13",
    "H1FS6"
  ),
  scoring_rule = c(
    "Clean all items; reverse-code positive affect items as 3 - item; sum all 19 items if complete",
    "Clean all items; sum the 4 screener items if complete",
    "Clean H1FS6 and keep as ordinal 0 to 3"
  ),
  valid_range = c(
    "0 to 57",
    "0 to 12",
    "0 to 3"
  ),
  note = c(
    "Binary version should be created later after analytic sample is locked",
    "Binary version should be created later after analytic sample is locked",
    "Binary version should be created later after analytic sample is locked"
  )
)

# Quality-check table
quality_checks <- tibble(
  variable_name = c(
    "cesd_full_score",
    "cesd_brief_score",
    "cesd_single_depressed_ord"
  ),
  n_total = nrow(addhealth_measures),
  n_nonmissing = c(
    sum(!is.na(addhealth_measures$cesd_full_score)),
    sum(!is.na(addhealth_measures$cesd_brief_score)),
    sum(!is.na(addhealth_measures$cesd_single_depressed_ord))
  ),
  n_missing = c(
    sum(is.na(addhealth_measures$cesd_full_score)),
    sum(is.na(addhealth_measures$cesd_brief_score)),
    sum(is.na(addhealth_measures$cesd_single_depressed_ord))
  ),
  min = c(
    min(addhealth_measures$cesd_full_score, na.rm = TRUE),
    min(addhealth_measures$cesd_brief_score, na.rm = TRUE),
    min(addhealth_measures$cesd_single_depressed_ord, na.rm = TRUE)
  ),
  max = c(
    max(addhealth_measures$cesd_full_score, na.rm = TRUE),
    max(addhealth_measures$cesd_brief_score, na.rm = TRUE),
    max(addhealth_measures$cesd_single_depressed_ord, na.rm = TRUE)
  ),
  mean = c(
    mean(addhealth_measures$cesd_full_score, na.rm = TRUE),
    mean(addhealth_measures$cesd_brief_score, na.rm = TRUE),
    mean(addhealth_measures$cesd_single_depressed_ord, na.rm = TRUE)
  ),
  sd = c(
    sd(addhealth_measures$cesd_full_score, na.rm = TRUE),
    sd(addhealth_measures$cesd_brief_score, na.rm = TRUE),
    sd(addhealth_measures$cesd_single_depressed_ord, na.rm = TRUE)
  )
)

# Reverse-code spot-check
reverse_code_spotcheck <- addhealth_measures %>%
  select(
    H1FS4, H1FS4_clean, H1FS4_rev,
    H1FS8, H1FS8_clean, H1FS8_rev,
    H1FS11, H1FS11_clean, H1FS11_rev,
    H1FS15, H1FS15_clean, H1FS15_rev
  ) %>%
  slice_head(n = 10)


################################################################################
# 6. Save outputs
################################################################################

# Save constructed dataset
saveRDS(
  addhealth_measures,
  "data/addhealth_depression_measures.rds"
)

# Save documentation and checks as one Excel workbook
write_xlsx(
  list(
    measure_documentation = measure_documentation,
    quality_checks = quality_checks,
    reverse_code_spotcheck = reverse_code_spotcheck
  ),
  "outputs/01_construct_measures_outputs.xlsx"
)


################################################################################
# 7. Print checks to console
################################################################################

quality_checks
reverse_code_spotcheck
```


# Step 2: Build analytic sample

```{r step-02}
################################################################################
# Script: 02_build_sample.R
# Purpose: Prepare final analytic sample for the Add Health P1 project
################################################################################



# Read the dataset created in 01_construct_measures_Rophence.R
addhealth_measures <- readRDS("data/addhealth_depression_measures.rds")

# Add a row ID so every respondent can be tracked
addhealth_measures <- addhealth_measures %>%
  mutate(respondent_row_id = row_number())


################################################################################
# 1. Define item lists
################################################################################

# Full modified CES-D items after cleaning/reverse-coding
full_score_items <- c(
  "H1FS1_clean",
  "H1FS2_clean",
  "H1FS3_clean",
  "H1FS4_rev",
  "H1FS5_clean",
  "H1FS6_clean",
  "H1FS7_clean",
  "H1FS8_rev",
  "H1FS9_clean",
  "H1FS10_clean",
  "H1FS11_rev",
  "H1FS12_clean",
  "H1FS13_clean",
  "H1FS14_clean",
  "H1FS15_rev",
  "H1FS16_clean",
  "H1FS17_clean",
  "H1FS18_clean",
  "H1FS19_clean"
)

# Brief screener items
brief_score_items <- c(
  "H1FS6_clean",
  "H1FS16_clean",
  "H1FS3_clean",
  "H1FS13_clean"
)

# Household roster relationship variables
roster_relation_items <- paste0("H1HR3", LETTERS[1:20])
roster_relation_code_items <- paste0(roster_relation_items, "_code")

# School connectedness items
school_connectedness_items <- c(
  "H1ED19",
  "H1ED20",
  "H1ED22",
  "H1ED24"
)


################################################################################
# 2. Helper cleaning functions
################################################################################

# Extract the actual Add Health numeric code from factor-labelled variables.
# Example: "(95) (95) 1995" becomes 95.
get_code <- function(x) {
  readr::parse_number(as.character(x))
}

clean_0_1 <- function(x) {
  x_num <- get_code(x)
  
  case_when(
    x_num %in% c(0, 1) ~ x_num,
    TRUE              ~ NA_real_
  )
}

clean_parent_education <- function(x) {
  x_num <- get_code(x)
  
  case_when(
    x_num %in% 1:10 ~ x_num,
    TRUE           ~ NA_real_
  )
}

clean_month <- function(x) {
  x_num <- get_code(x)
  
  case_when(
    x_num %in% 1:12 ~ x_num,
    TRUE           ~ NA_real_
  )
}

clean_birth_year <- function(x) {
  x_num <- get_code(x)
  
  case_when(
    x_num %in% c(96, 97, 98, 99, 996, 997, 998, 999) ~ NA_real_,
    x_num >= 1900 & x_num <= 2100                    ~ x_num,
    x_num >= 0 & x_num <= 95                         ~ 1900 + x_num,
    TRUE                                             ~ NA_real_
  )
}

clean_interview_year <- function(x) {
  x_num <- get_code(x)
  
  case_when(
    x_num %in% c(996, 997, 998, 999) ~ NA_real_,
    x_num >= 1900 & x_num <= 2100    ~ x_num,
    x_num >= 0 & x_num <= 99         ~ 1900 + x_num,
    TRUE                             ~ NA_real_
  )
}


################################################################################
# 3. Construct predictors and covariate
################################################################################

addhealth_vars <- addhealth_measures %>%
  
  ###########################################################################
# Gender, race source variables, parent education source variables, age parts
###########################################################################

mutate(
  gender = case_when(
    get_code(BIO_SEX) == 1 ~ "Male",
    get_code(BIO_SEX) == 2 ~ "Female",
    TRUE                  ~ NA_character_
  ),
  
  gender = factor(
    gender,
    levels = c("Male", "Female")
  ),
  
  hispanic_origin = clean_0_1(H1GI4),
  
  race_white = clean_0_1(H1GI6A),
  race_black = clean_0_1(H1GI6B),
  race_aian  = clean_0_1(H1GI6C),
  race_asian = clean_0_1(H1GI6D),
  race_other = clean_0_1(H1GI6E),
  
  H1RM1_clean = clean_parent_education(H1RM1),
  H1RF1_clean = clean_parent_education(H1RF1),
  
  H1PR4_code = get_code(H1PR4),
  
  interview_month = clean_month(IMONTH),
  interview_year  = clean_interview_year(IYEAR),
  
  birth_month = clean_month(H1GI1M),
  birth_year  = clean_birth_year(H1GI1Y)
) %>%
  
  ###########################################################################
# Race/ethnicity
###########################################################################

mutate(
  race_count_all = rowSums(
    select(
      .,
      race_white,
      race_black,
      race_aian,
      race_asian,
      race_other
    ) == 1,
    na.rm = TRUE
  ),
  
  race_count_no_aian = rowSums(
    select(
      .,
      race_white,
      race_black,
      race_asian,
      race_other
    ) == 1,
    na.rm = TRUE
  ),
  
  non_hispanic_or_missing = is.na(hispanic_origin) | hispanic_origin == 0,
  
  race_eth = case_when(
    hispanic_origin == 1 ~ "Hispanic",
    
    non_hispanic_or_missing & race_aian == 1 ~ "NH-AI/AN",
    
    non_hispanic_or_missing & race_count_all == 0 ~ NA_character_,
    
    non_hispanic_or_missing & race_count_no_aian == 1 & race_white == 1 ~
      "NH-White",
    
    non_hispanic_or_missing & race_count_no_aian == 1 & race_black == 1 ~
      "NH-Black",
    
    non_hispanic_or_missing & race_count_no_aian == 1 & race_asian == 1 ~
      "NH-Asian",
    
    non_hispanic_or_missing & race_count_no_aian == 1 & race_other == 1 ~
      "NH-Multiracial",
    
    non_hispanic_or_missing & race_count_no_aian >= 2 ~
      "NH-Multiracial",
    
    TRUE ~ NA_character_
  ),
  
  race_eth = factor(
    race_eth,
    levels = c(
      "NH-White",
      "NH-Black",
      "Hispanic",
      "NH-AI/AN",
      "NH-Asian",
      "NH-Multiracial"
    )
  )
) %>%
  
###########################################################################
# Highest resident parent education
###########################################################################

mutate(
  # Recode each parent's education into an ordered attainment category
  # before selecting the higher level. The original Add Health codes are
  # not numerically ordered by educational attainment.
  mother_education_n = case_when(
    H1RM1_clean %in% c(1, 2, 10) ~ 1,
    H1RM1_clean %in% c(3, 4, 5)  ~ 2,
    H1RM1_clean %in% c(6, 7)     ~ 3,
    H1RM1_clean %in% c(8, 9)     ~ 4,
    TRUE                         ~ NA_real_
  ),

  father_education_n = case_when(
    H1RF1_clean %in% c(1, 2, 10) ~ 1,
    H1RF1_clean %in% c(3, 4, 5)  ~ 2,
    H1RF1_clean %in% c(6, 7)     ~ 3,
    H1RF1_clean %in% c(8, 9)     ~ 4,
    TRUE                         ~ NA_real_
  ),

  # Use the higher educational-attainment category when both are available.
  # When only one parent has valid education data, use that parent's value.
  resident_parent_education_n = case_when(
    !is.na(mother_education_n) & !is.na(father_education_n) ~
      pmax(mother_education_n, father_education_n),

    !is.na(mother_education_n) ~ mother_education_n,

    !is.na(father_education_n) ~ father_education_n,

    TRUE ~ NA_real_
  ),

  resident_parent_education = factor(
    resident_parent_education_n,
    levels = 1:4,
    labels = c(
      "Less than high school",
      "HS/GED/vocational",
      "Some college/postsecondary vocational",
      "College degree or higher"
    )
  )
) %>%
  
  ###########################################################################
# Family structure from household roster
###########################################################################

mutate(
  across(
    all_of(roster_relation_items),
    get_code,
    .names = "{.col}_code"
  )
) %>%
  mutate(
    has_resident_father = rowSums(
      select(., all_of(roster_relation_code_items)) == 11,
      na.rm = TRUE
    ) > 0,
    
    has_resident_mother = rowSums(
      select(., all_of(roster_relation_code_items)) == 14,
      na.rm = TRUE
    ) > 0,
    
    family_structure = case_when(
      has_resident_mother & has_resident_father   ~ "Two parents",
      has_resident_mother & !has_resident_father  ~ "Mother only",
      !has_resident_mother & has_resident_father  ~ "Father only",
      !has_resident_mother & !has_resident_father ~ "Neither parent",
      TRUE ~ NA_character_
    ),
    
    family_structure = factor(
      family_structure,
      levels = c(
        "Two parents",
        "Mother only",
        "Father only",
        "Neither parent"
      )
    )
  ) %>%
  
  ###########################################################################
# School connectedness
###########################################################################

mutate(
  across(
    all_of(school_connectedness_items),
    ~ {
      x_num <- get_code(.x)
      
      case_when(
        x_num %in% 1:5        ~ x_num,
        x_num %in% c(6, 7, 8) ~ NA_real_,
        TRUE                  ~ NA_real_
      )
    },
    .names = "{.col}_clean"
  )
) %>%
  mutate(
    school_connectedness = rowMeans(
      select(
        .,
        H1ED19_clean,
        H1ED20_clean,
        H1ED22_clean,
        H1ED24_clean
      ),
      na.rm = TRUE
    ),
    
    school_connectedness = if_else(
      is.nan(school_connectedness),
      NA_real_,
      school_connectedness
    )
  ) %>%
  
  ###########################################################################
# Peer social support and age
###########################################################################

mutate(
  peer_support = case_when(
    H1PR4_code %in% 1:5          ~ H1PR4_code,
    H1PR4_code %in% c(6, 96, 98) ~ NA_real_,
    TRUE                         ~ NA_real_
  ),
  
  age = interview_year - birth_year +
    ((interview_month - birth_month) / 12),
  
  age = if_else(
    age >= 10 & age <= 25,
    age,
    NA_real_
  )
)


################################################################################
# 4. Age check before filtering
################################################################################

age_check <- addhealth_vars %>%
  summarise(
    n_total = n(),
    n_missing_interview_month = sum(is.na(interview_month)),
    n_missing_interview_year  = sum(is.na(interview_year)),
    n_missing_birth_month     = sum(is.na(birth_month)),
    n_missing_birth_year      = sum(is.na(birth_year)),
    n_missing_age             = sum(is.na(age)),
    min_age = min(age, na.rm = TRUE),
    max_age = max(age, na.rm = TRUE),
    mean_age = mean(age, na.rm = TRUE)
  )


################################################################################
# 5. Define variables required for the analytic sample
################################################################################

outcome_vars <- c(
  "cesd_full_score",
  "cesd_brief_score",
  "cesd_single_depressed_ord"
)

predictor_covariate_vars <- c(
  "gender",
  "race_eth",
  "resident_parent_education",
  "family_structure",
  "school_connectedness",
  "peer_support",
  "age"
)


################################################################################
# 6. Build sample flow documentation
################################################################################

add_sample_flow_row <- function(sample_flow, step_number, restriction,
                                before_n, after_n, original_n) {
  
  bind_rows(
    sample_flow,
    tibble(
      step_number = step_number,
      restriction = restriction,
      n_before = before_n,
      n_after = after_n,
      n_removed_at_step = before_n - after_n,
      percent_removed_at_step = if_else(
        before_n > 0,
        round(100 * (before_n - after_n) / before_n, 2),
        NA_real_
      ),
      percent_original_remaining = round(100 * after_n / original_n, 2),
      percent_original_removed_total = round(
        100 * (original_n - after_n) / original_n,
        2
      )
    )
  )
}

original_n <- nrow(addhealth_vars)

sample_flow <- tibble(
  step_number = 0,
  restriction = "Original dataset after 01_construct_measures_Rophence.R",
  n_before = original_n,
  n_after = original_n,
  n_removed_at_step = 0,
  percent_removed_at_step = 0,
  percent_original_remaining = 100,
  percent_original_removed_total = 0
)

current_data <- addhealth_vars


# Step 1: complete gender
before_n <- nrow(current_data)

current_data <- current_data %>%
  filter(!is.na(gender))

sample_flow <- add_sample_flow_row(
  sample_flow,
  1,
  "Complete gender: BIO_SEX coded male or female",
  before_n,
  nrow(current_data),
  original_n
)


# Step 2: complete race/ethnicity
before_n <- nrow(current_data)

current_data <- current_data %>%
  filter(!is.na(race_eth))

sample_flow <- add_sample_flow_row(
  sample_flow,
  2,
  "Complete constructed race/ethnicity",
  before_n,
  nrow(current_data),
  original_n
)


# Step 3: complete family structure
before_n <- nrow(current_data)

current_data <- current_data %>%
  filter(!is.na(family_structure))

sample_flow <- add_sample_flow_row(
  sample_flow,
  3,
  "Complete household-roster family structure",
  before_n,
  nrow(current_data),
  original_n
)


# Step 4: complete resident parent education
before_n <- nrow(current_data)

current_data <- current_data %>%
  filter(!is.na(resident_parent_education))

sample_flow <- add_sample_flow_row(
  sample_flow,
  4,
  "Complete highest resident parent education",
  before_n,
  nrow(current_data),
  original_n
)


# Step 5: complete school connectedness
before_n <- nrow(current_data)

current_data <- current_data %>%
  filter(!is.na(school_connectedness))

sample_flow <- add_sample_flow_row(
  sample_flow,
  5,
  "Complete school connectedness score",
  before_n,
  nrow(current_data),
  original_n
)


# Step 6: complete peer social support
before_n <- nrow(current_data)

current_data <- current_data %>%
  filter(!is.na(peer_support))

sample_flow <- add_sample_flow_row(
  sample_flow,
  6,
  "Complete peer social support: H1PR4",
  before_n,
  nrow(current_data),
  original_n
)


# Step 7: complete age
before_n <- nrow(current_data)

current_data <- current_data %>%
  filter(!is.na(age))

sample_flow <- add_sample_flow_row(
  sample_flow,
  7,
  "Complete computed age",
  before_n,
  nrow(current_data),
  original_n
)


# Step 8: complete full modified CES-D score
before_n <- nrow(current_data)

current_data <- current_data %>%
  filter(!is.na(cesd_full_score))

sample_flow <- add_sample_flow_row(
  sample_flow,
  8,
  "Complete full 19-item modified CES-D score",
  before_n,
  nrow(current_data),
  original_n
)


# Step 9: complete brief screener score
before_n <- nrow(current_data)

current_data <- current_data %>%
  filter(!is.na(cesd_brief_score))

sample_flow <- add_sample_flow_row(
  sample_flow,
  9,
  "Complete four-item brief screener score",
  before_n,
  nrow(current_data),
  original_n
)


# Step 10: complete single depressed mood item
before_n <- nrow(current_data)

current_data <- current_data %>%
  filter(!is.na(cesd_single_depressed_ord))

sample_flow <- add_sample_flow_row(
  sample_flow,
  10,
  "Complete single depressed mood item: H1FS6",
  before_n,
  nrow(current_data),
  original_n
)


################################################################################
# 7. Final analytic dataset
################################################################################

analytic_sample <- current_data %>%
  select(
    respondent_row_id,
    any_of(c("AID")),
    all_of(outcome_vars),
    all_of(predictor_covariate_vars),
    all_of(full_score_items)
  )


################################################################################
# 8. Quality checks
################################################################################

# Check 1: each outcome has the same final sample size
outcome_sample_check <- tibble(
  outcome_variable = outcome_vars,
  final_sample_size = nrow(analytic_sample),
  n_nonmissing = sapply(
    analytic_sample[outcome_vars],
    function(x) sum(!is.na(x))
  ),
  n_missing = sapply(
    analytic_sample[outcome_vars],
    function(x) sum(is.na(x))
  )
) %>%
  mutate(
    sample_identical_to_final_n = n_nonmissing == final_sample_size
  )


# Check 2: no missing values remain in predictors or covariates
predictor_missing_check <- analytic_sample %>%
  summarise(
    across(
      all_of(predictor_covariate_vars),
      ~ sum(is.na(.))
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable_name",
    values_to = "n_missing"
  ) %>%
  mutate(
    no_missing_values = n_missing == 0
  )


# Check 3: no missing values remain in outcomes
outcome_missing_check <- analytic_sample %>%
  summarise(
    across(
      all_of(outcome_vars),
      ~ sum(is.na(.))
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable_name",
    values_to = "n_missing"
  ) %>%
  mutate(
    no_missing_values = n_missing == 0
  )


# Check 4: 25% sample-loss decision rule
sample_loss_check <- tibble(
  original_n = original_n,
  final_n = nrow(analytic_sample),
  n_removed_total = original_n - nrow(analytic_sample),
  percent_removed_total = round(
    100 * (original_n - nrow(analytic_sample)) / original_n,
    2
  ),
  exceeds_25_percent_loss = percent_removed_total > 25,
  decision_rule_message = if_else(
    exceeds_25_percent_loss,
    "Flag for team: complete-case restriction removed more than 25% of original sample.",
    "Complete-case restriction did not exceed 25% sample-loss rule."
  )
)


# Check 5: final sample size to use in every later analysis
final_sample_size_check <- tibble(
  analytic_dataset = "data/addhealth_analytic_sample.rds",
  final_n = nrow(analytic_sample),
  note = "Use this same analytic dataset for all later descriptive, subgroup, regression, and classification analyses."
)


################################################################################
# 9. Save final dataset and documentation
################################################################################

# Save the final analytic dataset
saveRDS(
  analytic_sample,
  "data/addhealth_analytic_sample.rds"
)


# Save sample flow and quality checks in one Excel workbook
write_xlsx(
  list(
    age_check = age_check,
    sample_flow = sample_flow,
    sample_loss_check = sample_loss_check,
    outcome_sample_check = outcome_sample_check,
    predictor_missing_check = predictor_missing_check,
    outcome_missing_check = outcome_missing_check,
    final_sample_size_check = final_sample_size_check
  ),
  "outputs/02_build_sample_outputs.xlsx"
)


################################################################################
# 10. Print key checks to console
################################################################################

age_check
sample_flow
sample_loss_check
outcome_sample_check
predictor_missing_check
outcome_missing_check
final_sample_size_check
```


# Step 3 and Step 4: Descriptives, reliability, and classification thresholds

```{r step-03}
################################################################################
# Script: 03_descriptives.R
# Purpose: Produce descriptive statistics, distribution tables, reliability
#          estimates, and Figure 1 for the Add Health P1 project
################################################################################




################################################################################
# 1. Read analytic dataset
################################################################################

# This dataset was created in 02_build_sample.R.
# It should be the one used for every later analysis.
analytic_sample <- readRDS("data/addhealth_analytic_sample.rds")


################################################################################
# 2. Define outcome variables and item lists
################################################################################

outcome_vars <- c(
  "cesd_full_score",
  "cesd_brief_score",
  "cesd_single_depressed_ord"
)

full_score_items <- c(
  "H1FS1_clean",
  "H1FS2_clean",
  "H1FS3_clean",
  "H1FS4_rev",
  "H1FS5_clean",
  "H1FS6_clean",
  "H1FS7_clean",
  "H1FS8_rev",
  "H1FS9_clean",
  "H1FS10_clean",
  "H1FS11_rev",
  "H1FS12_clean",
  "H1FS13_clean",
  "H1FS14_clean",
  "H1FS15_rev",
  "H1FS16_clean",
  "H1FS17_clean",
  "H1FS18_clean",
  "H1FS19_clean"
)

brief_score_items <- c(
  "H1FS6_clean",
  "H1FS16_clean",
  "H1FS3_clean",
  "H1FS13_clean"
)


################################################################################
# 3. Basic sample-size check
################################################################################

sample_size_check <- tibble(
  analytic_n = nrow(analytic_sample),
  n_missing_full_score = sum(is.na(analytic_sample$cesd_full_score)),
  n_missing_brief_score = sum(is.na(analytic_sample$cesd_brief_score)),
  n_missing_single_item = sum(is.na(analytic_sample$cesd_single_depressed_ord))
)


################################################################################
# 4. Summary statistics for the three outcome measures
################################################################################

outcome_summary <- analytic_sample %>%
  summarise(
    full_n = sum(!is.na(cesd_full_score)),
    full_mean = mean(cesd_full_score, na.rm = TRUE),
    full_sd = sd(cesd_full_score, na.rm = TRUE),
    full_min = min(cesd_full_score, na.rm = TRUE),
    full_max = max(cesd_full_score, na.rm = TRUE),
    
    brief_n = sum(!is.na(cesd_brief_score)),
    brief_mean = mean(cesd_brief_score, na.rm = TRUE),
    brief_sd = sd(cesd_brief_score, na.rm = TRUE),
    brief_min = min(cesd_brief_score, na.rm = TRUE),
    brief_max = max(cesd_brief_score, na.rm = TRUE),
    
    single_n = sum(!is.na(cesd_single_depressed_ord)),
    single_mean = mean(cesd_single_depressed_ord, na.rm = TRUE),
    single_sd = sd(cesd_single_depressed_ord, na.rm = TRUE),
    single_min = min(cesd_single_depressed_ord, na.rm = TRUE),
    single_max = max(cesd_single_depressed_ord, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "statistic",
    values_to = "value"
  ) %>%
  separate(
    statistic,
    into = c("measure", "statistic"),
    sep = "_",
    extra = "merge"
  ) %>%
  pivot_wider(
    names_from = statistic,
    values_from = value
  ) %>%
  mutate(
    measure = case_when(
      measure == "full" ~ "Full 19-item modified CES-D",
      measure == "brief" ~ "Four-item brief screener",
      measure == "single" ~ "Single depressed-mood item",
      TRUE ~ measure
    ),
    possible_range = case_when(
      measure == "Full 19-item modified CES-D" ~ "0 to 57",
      measure == "Four-item brief screener" ~ "0 to 12",
      measure == "Single depressed-mood item" ~ "0 to 3",
      TRUE ~ NA_character_
    )
  ) %>%
  select(
    measure,
    n,
    mean,
    sd,
    min,
    max,
    possible_range
  )


################################################################################
# 5. Score distribution tables
################################################################################

# These tables show how many respondents fall at each observed score.
# This is useful for seeing floor effects, especially for the single item.

full_score_distribution <- analytic_sample %>%
  count(cesd_full_score, name = "n") %>%
  mutate(
    percent = round(100 * n / sum(n), 2),
    measure = "Full 19-item modified CES-D"
  ) %>%
  rename(score = cesd_full_score) %>%
  select(measure, score, n, percent)

brief_score_distribution <- analytic_sample %>%
  count(cesd_brief_score, name = "n") %>%
  mutate(
    percent = round(100 * n / sum(n), 2),
    measure = "Four-item brief screener"
  ) %>%
  rename(score = cesd_brief_score) %>%
  select(measure, score, n, percent)

single_item_distribution <- analytic_sample %>%
  count(cesd_single_depressed_ord, name = "n") %>%
  mutate(
    percent = round(100 * n / sum(n), 2),
    measure = "Single depressed-mood item"
  ) %>%
  rename(score = cesd_single_depressed_ord) %>%
  select(measure, score, n, percent)

score_distribution_all <- bind_rows(
  full_score_distribution,
  brief_score_distribution,
  single_item_distribution
)


################################################################################
# 6. Cronbach's alpha for full scale and brief screener
################################################################################

# psych::alpha() expects numeric item columns.
full_alpha_object <- psych::alpha(
  analytic_sample %>%
    select(all_of(full_score_items)),
  check.keys = FALSE
)

brief_alpha_object <- psych::alpha(
  analytic_sample %>%
    select(all_of(brief_score_items)),
  check.keys = FALSE
)

# Overall alpha values
reliability_summary <- tibble(
  measure = c(
    "Full 19-item modified CES-D",
    "Four-item brief screener"
  ),
  number_of_items = c(
    length(full_score_items),
    length(brief_score_items)
  ),
  cronbach_alpha = c(
    full_alpha_object$total$raw_alpha,
    brief_alpha_object$total$raw_alpha
  ),
  decision_rule = c(
    "If alpha < 0.70, investigate before proceeding",
    "If alpha < 0.60, document as limitation but proceed"
  ),
  decision_flag = c(
    if_else(
      full_alpha_object$total$raw_alpha < 0.70,
      "FLAG: full-scale alpha below 0.70",
      "OK: full-scale alpha is at least 0.70"
    ),
    if_else(
      brief_alpha_object$total$raw_alpha < 0.60,
      "NOTE: screener alpha below 0.60; document as limitation",
      "OK: screener alpha is at least 0.60"
    )
  )
)

# Alpha if each item is removed
full_alpha_drop <- full_alpha_object$alpha.drop %>%
  as.data.frame() %>%
  rownames_to_column("item") %>%
  as_tibble() %>%
  transmute(
    measure = "Full 19-item modified CES-D",
    item = item,
    overall_alpha = full_alpha_object$total$raw_alpha,
    alpha_if_item_deleted = raw_alpha,
    alpha_change_if_deleted = alpha_if_item_deleted - overall_alpha,
    item_flag = if_else(
      alpha_change_if_deleted > 0.02,
      "Alpha increases by more than 0.02 if removed; review item",
      "No large alpha increase if removed"
    )
  )

brief_alpha_drop <- brief_alpha_object$alpha.drop %>%
  as.data.frame() %>%
  rownames_to_column("item") %>%
  as_tibble() %>%
  transmute(
    measure = "Four-item brief screener",
    item = item,
    overall_alpha = brief_alpha_object$total$raw_alpha,
    alpha_if_item_deleted = raw_alpha,
    alpha_change_if_deleted = alpha_if_item_deleted - overall_alpha,
    item_flag = if_else(
      alpha_change_if_deleted > 0.02,
      "Alpha increases by more than 0.02 if removed; review item",
      "No large alpha increase if removed"
    )
  )

alpha_if_item_deleted <- bind_rows(
  full_alpha_drop,
  brief_alpha_drop
)


################################################################################
# 7. Figure 1: Distribution of depressive symptom measures
################################################################################

# Put the three outcomes into long format for plotting.
plot_data <- analytic_sample %>%
  select(
    cesd_full_score,
    cesd_brief_score,
    cesd_single_depressed_ord
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "measure",
    values_to = "score"
  ) %>%
  mutate(
    measure = case_when(
      measure == "cesd_full_score" ~ "Full 19-item modified CES-D",
      measure == "cesd_brief_score" ~ "Four-item brief screener",
      measure == "cesd_single_depressed_ord" ~ "Single depressed-mood item",
      TRUE ~ measure
    ),
    measure = factor(
      measure,
      levels = c(
        "Full 19-item modified CES-D",
        "Four-item brief screener",
        "Single depressed-mood item"
      )
    )
  )

# We use separate panels because the score ranges differ:
# full scale = 0 to 57, brief screener = 0 to 12, single item = 0 to 3.
figure1_distribution <- ggplot(
  plot_data,
  aes(x = score)
) +
  geom_histogram(
    binwidth = 1,
    boundary = -0.5,
    closed = "right"
  ) +
  facet_wrap(
    ~ measure,
    scales = "free_x",
    nrow = 1
  ) +
  labs(
    title = "Distribution of Depressive Symptom Measures",
    x = "Depressive symptom score",
    y = "Number of respondents") +
project_figure_theme

# Save Figure 1 as 300 DPI PNG
ggsave(
  filename = "outputs/figures/figure1_distribution_depressive_symptom_measures.png",
  plot = figure1_distribution,
  width = 12,
  height = 5,
  dpi = 300
)


################################################################################
# 8. Save tables to Excel workbook
################################################################################

write_xlsx(
  list(
    sample_size_check = sample_size_check,
    outcome_summary = outcome_summary,
    score_distribution_all = score_distribution_all,
    full_score_distribution = full_score_distribution,
    brief_score_distribution = brief_score_distribution,
    single_item_distribution = single_item_distribution,
    reliability_summary = reliability_summary,
    alpha_if_item_deleted = alpha_if_item_deleted
  ),
  "outputs/03_descriptives_outputs.xlsx"
)


################################################################################
# 9. Print key outputs to console
################################################################################

sample_size_check
outcome_summary
reliability_summary
alpha_if_item_deleted
figure1_distribution


################################################################################
# 10. Step 4: Apply 80th-percentile classification thresholds
################################################################################

# The binary elevated-distress variables are created only after the analytic
# sample is locked. This ensures that all three measures use the same respondents.
#
# Decision rule:
# - Use the 80th percentile of each measure's distribution in the locked analytic
#   sample.
# - Use an empirical percentile so the cutoff is an observed score.
# - Include all respondents tied at the cutoff in the elevated-distress group.

get_empirical_80th_percentile <- function(x) {
  as.numeric(
    quantile(
      x,
      probs = 0.80,
      na.rm = TRUE,
      type = 1,
      names = FALSE
    )
  )
}

# Calculate the locked 80th-percentile cutoffs
full_cutoff <- get_empirical_80th_percentile(
  analytic_sample$cesd_full_score
)

brief_cutoff <- get_empirical_80th_percentile(
  analytic_sample$cesd_brief_score
)

single_cutoff <- get_empirical_80th_percentile(
  analytic_sample$cesd_single_depressed_ord
)


# Create binary elevated-distress variables
# 1 = elevated distress
# 0 = not elevated distress
analytic_sample_thresholded <- analytic_sample %>%
  mutate(
    cesd_full_elevated = if_else(
      cesd_full_score >= full_cutoff,
      1L,
      0L
    ),
    
    cesd_brief_elevated = if_else(
      cesd_brief_score >= brief_cutoff,
      1L,
      0L
    ),
    
    cesd_single_depressed_elevated = if_else(
      cesd_single_depressed_ord >= single_cutoff,
      1L,
      0L
    ),
    
    n_elevated_measures =
      cesd_full_elevated +
      cesd_brief_elevated +
      cesd_single_depressed_elevated,
    
    classification_overlap = case_when(
      n_elevated_measures == 0 ~ "Not elevated on any measure",
      n_elevated_measures == 1 ~ "Elevated on one measure",
      n_elevated_measures == 2 ~ "Elevated on two measures",
      n_elevated_measures == 3 ~ "Elevated on all three measures",
      TRUE ~ NA_character_
    ),
    
    classification_overlap = factor(
      classification_overlap,
      levels = c(
        "Not elevated on any measure",
        "Elevated on one measure",
        "Elevated on two measures",
        "Elevated on all three measures"
      )
    )
  )


################################################################################
# 11. Threshold documentation
################################################################################

make_threshold_row <- function(data, measure_label, score_variable,
                               binary_variable, cutoff_score) {
  
  tibble(
    measure = measure_label,
    score_variable = score_variable,
    binary_variable = binary_variable,
    percentile_rule = "80th percentile within locked analytic sample",
    quantile_rule = "Empirical quantile using type = 1",
    cutoff_score = cutoff_score,
    elevated_definition = paste0(score_variable, " >= ", cutoff_score),
    n_total = nrow(data),
    n_at_cutoff = sum(data[[score_variable]] == cutoff_score),
    n_elevated = sum(data[[binary_variable]] == 1),
    percent_elevated = round(
      100 * sum(data[[binary_variable]] == 1) / nrow(data),
      2
    ),
    tie_rule = "All respondents with scores equal to the cutoff are included as elevated."
  )
}

threshold_summary <- bind_rows(
  make_threshold_row(
    analytic_sample_thresholded,
    "Full 19-item modified CES-D",
    "cesd_full_score",
    "cesd_full_elevated",
    full_cutoff
  ),
  
  make_threshold_row(
    analytic_sample_thresholded,
    "Four-item brief screener",
    "cesd_brief_score",
    "cesd_brief_elevated",
    brief_cutoff
  ),
  
  make_threshold_row(
    analytic_sample_thresholded,
    "Single depressed-mood item",
    "cesd_single_depressed_ord",
    "cesd_single_depressed_elevated",
    single_cutoff
  )
)


################################################################################
# 12. Classification prevalence and overlap tables
################################################################################

classification_prevalence <- analytic_sample_thresholded %>%
  summarise(
    full_n_elevated = sum(cesd_full_elevated == 1),
    full_percent_elevated = round(100 * mean(cesd_full_elevated == 1), 2),
    
    brief_n_elevated = sum(cesd_brief_elevated == 1),
    brief_percent_elevated = round(100 * mean(cesd_brief_elevated == 1), 2),
    
    single_n_elevated = sum(cesd_single_depressed_elevated == 1),
    single_percent_elevated = round(
      100 * mean(cesd_single_depressed_elevated == 1),
      2
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "statistic",
    values_to = "value"
  ) %>%
  separate(
    statistic,
    into = c("measure", "statistic"),
    sep = "_",
    extra = "merge"
  ) %>%
  pivot_wider(
    names_from = statistic,
    values_from = value
  ) %>%
  mutate(
    measure = case_when(
      measure == "full" ~ "Full 19-item modified CES-D",
      measure == "brief" ~ "Four-item brief screener",
      measure == "single" ~ "Single depressed-mood item",
      TRUE ~ measure
    )
  ) %>%
  rename(
    n_elevated = n_elevated,
    percent_elevated = percent_elevated
  )

classification_overlap_table <- analytic_sample_thresholded %>%
  count(classification_overlap, name = "n") %>%
  mutate(
    percent = round(100 * n / sum(n), 2)
  )


################################################################################
# 13. Quality checks for binary classification variables
################################################################################

binary_vars <- c(
  "cesd_full_elevated",
  "cesd_brief_elevated",
  "cesd_single_depressed_elevated"
)

binary_missing_check <- analytic_sample_thresholded %>%
  summarise(
    across(
      all_of(binary_vars),
      ~ sum(is.na(.))
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "binary_variable",
    values_to = "n_missing"
  ) %>%
  mutate(
    no_missing_values = n_missing == 0
  )

binary_value_check <- analytic_sample_thresholded %>%
  summarise(
    across(
      all_of(binary_vars),
      ~ paste(sort(unique(.)), collapse = ", ")
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "binary_variable",
    values_to = "observed_values"
  ) %>%
  mutate(
    expected_values = "0, 1",
    values_correct = observed_values == expected_values
  )

threshold_sample_size_check <- tibble(
  analytic_sample_n = nrow(analytic_sample),
  thresholded_sample_n = nrow(analytic_sample_thresholded),
  same_n = analytic_sample_n == thresholded_sample_n,
  note = "Thresholding should not add or remove respondents."
)


################################################################################
# 14. Save thresholded analytic dataset and threshold outputs
################################################################################

# Save analytic dataset with binary elevated-distress variables.
# Later scripts should read this file.
saveRDS(
  analytic_sample_thresholded,
  "data/addhealth_analytic_sample_thresholded.rds"
)

# Save threshold documentation and checks.
write_xlsx(
  list(
    threshold_summary = threshold_summary,
    classification_prevalence = classification_prevalence,
    classification_overlap_table = classification_overlap_table,
    binary_missing_check = binary_missing_check,
    binary_value_check = binary_value_check,
    threshold_sample_size_check = threshold_sample_size_check
  ),
  "outputs/03_classification_threshold_outputs.xlsx"
)


################################################################################
# 15. Print Step 4 outputs to console
################################################################################

threshold_summary
classification_prevalence
classification_overlap_table
binary_missing_check
binary_value_check
threshold_sample_size_check
```


# Step 5: Subgroup comparisons

```{r step-04}
################################################################################
# Script: 04_subgroups.R
# Purpose: Produce subgroup comparisons by gender and race/ethnicity for the
#          Add Health P1 depression measurement comparison project
################################################################################




################################################################################
# 1. Read thresholded analytic dataset
################################################################################

# This dataset should come from 03_descriptives.R after Step 4 thresholds
# have been applied.
analytic_sample <- readRDS("data/addhealth_analytic_sample_thresholded.rds")


################################################################################
# 2. Define continuous and binary outcome variables
################################################################################

continuous_outcomes <- c(
  "cesd_full_score",
  "cesd_brief_score",
  "cesd_single_depressed_ord"
)

binary_outcomes <- c(
  "cesd_full_elevated",
  "cesd_brief_elevated",
  "cesd_single_depressed_elevated"
)


################################################################################
# 3. Basic quality checks
################################################################################

required_vars <- c(
  "gender",
  "race_eth",
  continuous_outcomes,
  binary_outcomes
)

missing_required_vars <- setdiff(required_vars, names(analytic_sample))

if (length(missing_required_vars) > 0) {
  stop(
    paste(
      "These required variables are missing from the dataset:",
      paste(missing_required_vars, collapse = ", ")
    )
  )
}

sample_size_check <- tibble(
  analytic_n = nrow(analytic_sample),
  n_missing_gender = sum(is.na(analytic_sample$gender)),
  n_missing_race_eth = sum(is.na(analytic_sample$race_eth)),
  n_missing_full_score = sum(is.na(analytic_sample$cesd_full_score)),
  n_missing_brief_score = sum(is.na(analytic_sample$cesd_brief_score)),
  n_missing_single_item = sum(is.na(analytic_sample$cesd_single_depressed_ord)),
  n_missing_full_binary = sum(is.na(analytic_sample$cesd_full_elevated)),
  n_missing_brief_binary = sum(is.na(analytic_sample$cesd_brief_elevated)),
  n_missing_single_binary = sum(is.na(analytic_sample$cesd_single_depressed_elevated))
)


################################################################################
# 4. Helper functions for subgroup summaries
################################################################################

label_measure <- function(x) {
  case_when(
    x == "cesd_full_score" ~ "Full 19-item modified CES-D",
    x == "cesd_brief_score" ~ "Four-item brief screener",
    x == "cesd_single_depressed_ord" ~ "Single depressed-mood item",
    x == "cesd_full_elevated" ~ "Full 19-item modified CES-D",
    x == "cesd_brief_elevated" ~ "Four-item brief screener",
    x == "cesd_single_depressed_elevated" ~ "Single depressed-mood item",
    TRUE ~ x
  )
}


make_continuous_subgroup_summary <- function(data, subgroup_var, subgroup_label) {
  
  data %>%
    select(
      all_of(subgroup_var),
      all_of(continuous_outcomes)
    ) %>%
    pivot_longer(
      cols = all_of(continuous_outcomes),
      names_to = "outcome_variable",
      values_to = "score"
    ) %>%
    mutate(
      subgroup_variable = subgroup_label,
      subgroup_level = as.character(.data[[subgroup_var]]),
      measure = label_measure(outcome_variable)
    ) %>%
    group_by(
      subgroup_variable,
      subgroup_level,
      outcome_variable,
      measure
    ) %>%
    summarise(
      n = n(),
      mean_score = mean(score),
      sd_score = sd(score),
      min_score = min(score),
      max_score = max(score),
      .groups = "drop"
    ) %>%
    mutate(
      mean_score = round(mean_score, 3),
      sd_score = round(sd_score, 3)
    )
}


make_binary_subgroup_summary <- function(data, subgroup_var, subgroup_label) {
  
  data %>%
    select(
      all_of(subgroup_var),
      all_of(binary_outcomes)
    ) %>%
    pivot_longer(
      cols = all_of(binary_outcomes),
      names_to = "binary_variable",
      values_to = "elevated"
    ) %>%
    mutate(
      subgroup_variable = subgroup_label,
      subgroup_level = as.character(.data[[subgroup_var]]),
      measure = label_measure(binary_variable)
    ) %>%
    group_by(
      subgroup_variable,
      subgroup_level,
      binary_variable,
      measure
    ) %>%
    summarise(
      n = n(),
      n_elevated = sum(elevated == 1),
      prevalence_percent = 100 * mean(elevated == 1),
      .groups = "drop"
    ) %>%
    mutate(
      prevalence_percent = round(prevalence_percent, 2)
    )
}


################################################################################
# 5. Mean continuous scores by gender and race/ethnicity
################################################################################

gender_continuous_summary <- make_continuous_subgroup_summary(
  data = analytic_sample,
  subgroup_var = "gender",
  subgroup_label = "Gender"
)

race_continuous_summary <- make_continuous_subgroup_summary(
  data = analytic_sample,
  subgroup_var = "race_eth",
  subgroup_label = "Race/ethnicity"
)

continuous_subgroup_summary <- bind_rows(
  gender_continuous_summary,
  race_continuous_summary
)


################################################################################
# 6. Elevated-distress prevalence by gender and race/ethnicity
################################################################################

gender_prevalence_summary <- make_binary_subgroup_summary(
  data = analytic_sample,
  subgroup_var = "gender",
  subgroup_label = "Gender"
)

race_prevalence_summary <- make_binary_subgroup_summary(
  data = analytic_sample,
  subgroup_var = "race_eth",
  subgroup_label = "Race/ethnicity"
)

prevalence_subgroup_summary <- bind_rows(
  gender_prevalence_summary,
  race_prevalence_summary
)


################################################################################
# 7. Optional disparity summary tables
################################################################################

# These tables summarize the spread across subgroup levels.
# They are useful for quickly seeing whether subgroup gaps look larger
# under one measurement approach than another.

continuous_disparity_summary <- continuous_subgroup_summary %>%
  group_by(
    subgroup_variable,
    outcome_variable,
    measure
  ) %>%
  summarise(
    lowest_mean_score = min(mean_score),
    highest_mean_score = max(mean_score),
    mean_score_range = highest_mean_score - lowest_mean_score,
    subgroup_with_lowest_mean = subgroup_level[which.min(mean_score)],
    subgroup_with_highest_mean = subgroup_level[which.max(mean_score)],
    .groups = "drop"
  ) %>%
  mutate(
    mean_score_range = round(mean_score_range, 3)
  )

prevalence_disparity_summary <- prevalence_subgroup_summary %>%
  group_by(
    subgroup_variable,
    binary_variable,
    measure
  ) %>%
  summarise(
    lowest_prevalence_percent = min(prevalence_percent),
    highest_prevalence_percent = max(prevalence_percent),
    prevalence_range_percentage_points =
      highest_prevalence_percent - lowest_prevalence_percent,
    subgroup_with_lowest_prevalence = subgroup_level[which.min(prevalence_percent)],
    subgroup_with_highest_prevalence = subgroup_level[which.max(prevalence_percent)],
    .groups = "drop"
  ) %>%
  mutate(
    prevalence_range_percentage_points =
      round(prevalence_range_percentage_points, 2)
  )


################################################################################
# 8. Figure 3A: Mean continuous scores by subgroup
################################################################################

# Note:
# The three continuous measures have different score ranges:
# full scale = 0 to 57
# brief screener = 0 to 12
# single item = 0 to 3
# For this reason, the figure uses separate rows for each measure.

# Create cleaner labels for plotting
continuous_subgroup_summary_plot <- continuous_subgroup_summary %>%
  mutate(
    mean_label = round(mean_score, 2)
  )

figure3a_mean_scores <- ggplot(
  continuous_subgroup_summary_plot,
  aes(
    x = subgroup_level,
    y = mean_score,
    fill = measure
  )
) +
  geom_text(
  aes(label = mean_label),
  vjust = -0.15,
  size = 4,
  fontface = "bold",
  colour = "black"
) +
  geom_col(
    show.legend = FALSE,
    width = 0.55
  ) +
facet_grid(
  rows = vars(measure),
  cols = vars(subgroup_variable),
  scales = "free",
  space = "free_x",
  labeller = labeller(
    measure = c(
      "Full 19-item modified CES-D" = "Full 19-item CES-D",
      "Four-item brief screener" = "Four-item screener",
      "Single depressed-mood item" = "Single item"
    )
  )
) +
  scale_x_discrete(drop = TRUE) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.12))
  ) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Mean Depressive Symptom Scores by Demographic Subgroup",
    x = NULL,
    y = "Mean score") +
project_figure_theme +
theme(
  axis.text.x = element_text(
    angle = 0,
    hjust = 0.5,
    face = "bold",
    colour = "black"
  )
)

ggsave(
  filename = "outputs/figures/figure3a_mean_scores_by_subgroup.png",
  plot = figure3a_mean_scores,
  width = 12,
  height = 7,
  dpi = 300
)


################################################################################
# 9. Figure 3B: Elevated-distress prevalence by subgroup
################################################################################

# Create a shorter label just for plotting
prevalence_subgroup_summary_plot <- prevalence_subgroup_summary %>%
  mutate(
    prevalence_label = paste0(round(prevalence_percent, 1), "%")
  )

figure3b_prevalence <- ggplot(
  prevalence_subgroup_summary_plot,
  aes(
    x = subgroup_level,
    y = prevalence_percent,
    fill = measure
  )
) +
  geom_text(
  aes(label = prevalence_label),
  position = position_dodge(width = 0.8),
  vjust = -0.25,
  size = 3.5,
  fontface = "bold",
  colour = "black"
) +
  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.7
  ) +
  facet_grid(
    . ~ subgroup_variable,
    scales = "free_x",
    space = "free_x"
  ) +
  scale_x_discrete(drop = TRUE) +
  labs(
    title = "Elevated-Distress Prevalence by Demographic Subgroup",
    x = NULL,
    y = "Percent classified as elevated distress",
    fill = "Measurement approach") +
  coord_cartesian(
    ylim = c(0, max(prevalence_subgroup_summary_plot$prevalence_percent) + 10)
  ) +
project_figure_theme +
theme(
  axis.text.x = element_text(
    angle = 0,
    hjust = 0.5,
    face = "bold",
    colour = "black"
  )
)

ggsave(
  filename = "outputs/figures/figure3b_prevalence_by_subgroup.png",
  plot = figure3b_prevalence,
  width = 12,
  height = 6,
  dpi = 300
)


################################################################################
# 10. Save subgroup tables
################################################################################

write_xlsx(
  list(
    sample_size_check = sample_size_check,
    continuous_subgroup_summary = continuous_subgroup_summary,
    prevalence_subgroup_summary = prevalence_subgroup_summary,
    continuous_disparity_summary = continuous_disparity_summary,
    prevalence_disparity_summary = prevalence_disparity_summary
  ),
  "outputs/04_subgroup_comparisons_outputs.xlsx"
)


################################################################################
# 11. Print key outputs to console
################################################################################

sample_size_check
continuous_subgroup_summary
prevalence_subgroup_summary
continuous_disparity_summary
prevalence_disparity_summary
figure3a_mean_scores
figure3b_prevalence
```


# Step 6: Regression models

```{r step-05}
################################################################################
# Script: 05_regressions.R
# Purpose: Run linear and logistic regression models for the Add Health P1
#          depression measurement comparison project
################################################################################




################################################################################
# 1. Read thresholded analytic dataset
################################################################################

analytic_sample <- readRDS("data/addhealth_analytic_sample_thresholded.rds")


################################################################################
# 2. Define outcomes and predictors
################################################################################

continuous_outcomes <- c(
  "cesd_full_score",
  "cesd_brief_score",
  "cesd_single_depressed_ord"
)

binary_outcomes <- c(
  "cesd_full_elevated",
  "cesd_brief_elevated",
  "cesd_single_depressed_elevated"
)

predictor_vars <- c(
  "gender",
  "race_eth",
  "resident_parent_education",
  "family_structure",
  "school_connectedness",
  "peer_support",
  "age"
)

categorical_predictors <- c(
  "gender",
  "race_eth",
  "resident_parent_education",
  "family_structure"
)

continuous_predictors <- c(
  "school_connectedness",
  "peer_support",
  "age"
)


################################################################################
# 3. Basic checks before modeling
################################################################################

required_vars <- c(
  continuous_outcomes,
  binary_outcomes,
  predictor_vars
)

missing_required_vars <- setdiff(required_vars, names(analytic_sample))

if (length(missing_required_vars) > 0) {
  stop(
    paste(
      "These required variables are missing:",
      paste(missing_required_vars, collapse = ", ")
    )
  )
}

regression_sample_check <- tibble(
  analytic_n = nrow(analytic_sample),
  n_missing_any_model_variable = analytic_sample %>%
    select(all_of(required_vars)) %>%
    filter(if_any(everything(), is.na)) %>%
    nrow(),
  no_missing_model_variables = n_missing_any_model_variable == 0
)

binary_outcome_check <- analytic_sample %>%
  summarise(
    across(
      all_of(binary_outcomes),
      ~ paste(sort(unique(.)), collapse = ", ")
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "binary_outcome",
    values_to = "observed_values"
  ) %>%
  mutate(
    expected_values = "0, 1",
    values_correct = observed_values == expected_values
  )


################################################################################
# 4. Set reference categories
################################################################################

analytic_sample <- analytic_sample %>%
  mutate(
    gender = factor(
      gender,
      levels = c("Male", "Female")
    ),
    
    race_eth = factor(
      race_eth,
      levels = c(
        "NH-White",
        "NH-Black",
        "Hispanic",
        "NH-AI/AN",
        "NH-Asian",
        "NH-Multiracial"
      )
    ),
    
    resident_parent_education = forcats::fct_relevel(
      factor(resident_parent_education),
      "Less than high school"
    ),
    
    family_structure = factor(
      family_structure,
      levels = c(
        "Two parents",
        "Mother only",
        "Father only",
        "Neither parent"
      )
    )
  )

reference_level_table <- tibble(
  variable = c(
    "gender",
    "race_eth",
    "resident_parent_education",
    "family_structure"
  ),
  reference_level = c(
    levels(analytic_sample$gender)[1],
    levels(analytic_sample$race_eth)[1],
    levels(analytic_sample$resident_parent_education)[1],
    levels(analytic_sample$family_structure)[1]
  )
)


################################################################################
# 5. Create standardized variables for continuous linear models
################################################################################

# For standardized continuous models:
# - continuous outcomes are standardized
# - continuous predictors are standardized
# - categorical predictors stay as factors
#
# This lets coefficients be compared across the three outcome scales.
# For categorical predictors, coefficients are interpreted as SD-unit outcome
# differences relative to the reference group.

analytic_sample <- analytic_sample %>%
  mutate(
    cesd_full_score_z = as.numeric(scale(cesd_full_score)),
    cesd_brief_score_z = as.numeric(scale(cesd_brief_score)),
    cesd_single_depressed_ord_z = as.numeric(scale(cesd_single_depressed_ord)),
    
    school_connectedness_z = as.numeric(scale(school_connectedness)),
    peer_support_z = as.numeric(scale(peer_support)),
    age_z = as.numeric(scale(age))
  )

standardized_outcomes <- c(
  "cesd_full_score_z",
  "cesd_brief_score_z",
  "cesd_single_depressed_ord_z"
)

standardized_predictor_vars <- c(
  "gender",
  "race_eth",
  "resident_parent_education",
  "family_structure",
  "school_connectedness_z",
  "peer_support_z",
  "age_z"
)


################################################################################
# 6. Helper functions
################################################################################

measure_label <- function(x) {
  case_when(
    x %in% c("cesd_full_score", "cesd_full_score_z", "cesd_full_elevated") ~
      "Full 19-item modified CES-D",
    
    x %in% c("cesd_brief_score", "cesd_brief_score_z", "cesd_brief_elevated") ~
      "Four-item brief screener",
    
    x %in% c(
      "cesd_single_depressed_ord",
      "cesd_single_depressed_ord_z",
      "cesd_single_depressed_elevated"
    ) ~
      "Single depressed-mood item",
    
    TRUE ~ x
  )
}


clean_term_label <- function(term) {
  case_when(
    term == "(Intercept)" ~ "Intercept",
    
    term == "genderFemale" ~
      "Gender: Female vs Male",
    
    str_starts(term, "race_eth") ~
      paste0(
        "Race/ethnicity: ",
        str_remove(term, "^race_eth"),
        " vs NH-White"
      ),
    
    str_starts(term, "resident_parent_education") ~
      paste0(
        "Resident parent education: ",
        str_remove(term, "^resident_parent_education"),
        " vs less than high school"
      ),
    
    str_starts(term, "family_structure") ~
      paste0(
        "Family structure: ",
        str_remove(term, "^family_structure"),
        " vs two parents"
      ),
    
    term == "school_connectedness" ~
      "School connectedness",
    
    term == "peer_support" ~
      "Peer social support",
    
    term == "age" ~
      "Age",
    
    term == "school_connectedness_z" ~
      "School connectedness, per SD",
    
    term == "peer_support_z" ~
      "Peer social support, per SD",
    
    term == "age_z" ~
      "Age, per SD",
    
    TRUE ~ term
  )
}


make_formula <- function(outcome, predictors) {
  as.formula(
    paste(
      outcome,
      "~",
      paste(predictors, collapse = " + ")
    )
  )
}


tidy_lm_wald <- function(model, outcome_name, model_type, coefficient_scale) {
  broom::tidy(model) %>%
    mutate(
      conf.low = estimate - 1.96 * std.error,
      conf.high = estimate + 1.96 * std.error,
      outcome = outcome_name,
      measure = measure_label(outcome_name),
      model_type = model_type,
      coefficient_scale = coefficient_scale,
      term_label = clean_term_label(term)
    ) %>%
    select(
      outcome,
      measure,
      model_type,
      coefficient_scale,
      term,
      term_label,
      estimate,
      std.error,
      conf.low,
      conf.high,
      statistic,
      p.value
    )
}


tidy_glm_or_wald <- function(model, outcome_name) {
  broom::tidy(model) %>%
    mutate(
      log_odds_estimate = estimate,
      log_odds_std_error = std.error,
      log_odds_conf_low = estimate - 1.96 * std.error,
      log_odds_conf_high = estimate + 1.96 * std.error,
      
      odds_ratio = exp(log_odds_estimate),
      conf.low = exp(log_odds_conf_low),
      conf.high = exp(log_odds_conf_high),
      
      outcome = outcome_name,
      measure = measure_label(outcome_name),
      model_type = "Logistic regression",
      coefficient_scale = "Odds ratio",
      term_label = clean_term_label(term)
    ) %>%
    select(
      outcome,
      measure,
      model_type,
      coefficient_scale,
      term,
      term_label,
      log_odds_estimate,
      log_odds_std_error,
      odds_ratio,
      conf.low,
      conf.high,
      statistic,
      p.value
    )
}


fit_glm_with_warnings <- function(formula, data) {
  
  warning_messages <- character(0)
  
  model <- withCallingHandlers(
    glm(
      formula = formula,
      data = data,
      family = binomial()
    ),
    warning = function(w) {
      warning_messages <<- c(warning_messages, conditionMessage(w))
      invokeRestart("muffleWarning")
    }
  )
  
  list(
    model = model,
    warnings = unique(warning_messages)
  )
}


calculate_vif_from_model <- function(model) {
  
  x_matrix <- model.matrix(model)
  
  x_matrix <- x_matrix[
    ,
    colnames(x_matrix) != "(Intercept)",
    drop = FALSE
  ]
  
  if (ncol(x_matrix) < 2) {
    return(
      tibble(
        model_term = colnames(x_matrix),
        vif = NA_real_,
        vif_flag = "Not enough predictors to calculate VIF"
      )
    )
  }
  
  vif_values <- map_dbl(
    seq_len(ncol(x_matrix)),
    function(j) {
      
      x_j <- x_matrix[, j]
      x_others <- x_matrix[, -j, drop = FALSE]
      
      r_squared <- summary(
        lm(x_j ~ x_others)
      )$r.squared
      
      1 / (1 - r_squared)
    }
  )
  
  tibble(
    model_term = colnames(x_matrix),
    vif = vif_values,
    vif_flag = case_when(
      is.infinite(vif) ~ "FLAG: infinite VIF",
      vif > 5 ~ "FLAG: VIF above 5",
      TRUE ~ "OK"
    )
  )
}


check_categorical_separation <- function(data, binary_outcome, categorical_vars) {
  
  map_dfr(
    categorical_vars,
    function(var_name) {
      
      data %>%
        group_by(
          predictor = .data[[var_name]],
          .drop = FALSE
        ) %>%
        summarise(
          n = n(),
          n_events = sum(.data[[binary_outcome]] == 1),
          n_nonevents = sum(.data[[binary_outcome]] == 0),
          event_rate = mean(.data[[binary_outcome]] == 1),
          .groups = "drop"
        ) %>%
        mutate(
          outcome = binary_outcome,
          measure = measure_label(binary_outcome),
          predictor_variable = var_name,
          separation_flag = case_when(
            n_events == 0 ~ "FLAG: zero events in this category",
            n_nonevents == 0 ~ "FLAG: zero non-events in this category",
            TRUE ~ "OK"
          )
        ) %>%
        select(
          outcome,
          measure,
          predictor_variable,
          predictor,
          n,
          n_events,
          n_nonevents,
          event_rate,
          separation_flag
        )
    }
  )
}


################################################################################
# 7. Fit unstandardized linear regression models
################################################################################

linear_unstandardized_models <- map(
  continuous_outcomes,
  function(y) {
    lm(
      formula = make_formula(y, predictor_vars),
      data = analytic_sample
    )
  }
)

names(linear_unstandardized_models) <- continuous_outcomes

linear_unstandardized_results <- map_dfr(
  continuous_outcomes,
  function(y) {
    tidy_lm_wald(
      model = linear_unstandardized_models[[y]],
      outcome_name = y,
      model_type = "Linear regression",
      coefficient_scale = "Unstandardized coefficient"
    )
  }
)


################################################################################
# 8. Fit standardized linear regression models
################################################################################

linear_standardized_models <- map(
  standardized_outcomes,
  function(y) {
    lm(
      formula = make_formula(y, standardized_predictor_vars),
      data = analytic_sample
    )
  }
)

names(linear_standardized_models) <- standardized_outcomes

linear_standardized_results <- map_dfr(
  standardized_outcomes,
  function(y) {
    tidy_lm_wald(
      model = linear_standardized_models[[y]],
      outcome_name = y,
      model_type = "Linear regression",
      coefficient_scale = "Standardized coefficient"
    )
  }
)


################################################################################
# 9. Fit logistic regression models
################################################################################

logistic_fit_objects <- map(
  binary_outcomes,
  function(y) {
    fit_glm_with_warnings(
      formula = make_formula(y, predictor_vars),
      data = analytic_sample
    )
  }
)

names(logistic_fit_objects) <- binary_outcomes

logistic_models <- map(
  logistic_fit_objects,
  "model"
)

logistic_results <- map_dfr(
  binary_outcomes,
  function(y) {
    tidy_glm_or_wald(
      model = logistic_models[[y]],
      outcome_name = y
    )
  }
)


################################################################################
# 10. Model fit summaries
################################################################################

linear_unstandardized_fit_summary <- map_dfr(
  continuous_outcomes,
  function(y) {
    broom::glance(linear_unstandardized_models[[y]]) %>%
      mutate(
        outcome = y,
        measure = measure_label(y),
        model_type = "Linear regression",
        coefficient_scale = "Unstandardized"
      )
  }
)

linear_standardized_fit_summary <- map_dfr(
  standardized_outcomes,
  function(y) {
    broom::glance(linear_standardized_models[[y]]) %>%
      mutate(
        outcome = y,
        measure = measure_label(y),
        model_type = "Linear regression",
        coefficient_scale = "Standardized"
      )
  }
)

logistic_fit_summary <- map_dfr(
  binary_outcomes,
  function(y) {
    broom::glance(logistic_models[[y]]) %>%
      mutate(
        outcome = y,
        measure = measure_label(y),
        model_type = "Logistic regression",
        coefficient_scale = "Odds ratio",
        converged = logistic_models[[y]]$converged
      )
  }
)

model_fit_summary <- bind_rows(
  linear_unstandardized_fit_summary,
  linear_standardized_fit_summary,
  logistic_fit_summary
)


################################################################################
# 11. Multicollinearity diagnostics
################################################################################

# VIF does not depend on the outcome when the predictor set is the same.
# We calculate it using the full-scale unstandardized linear model.
# This is a column-level VIF based on the model matrix, so categorical
# predictors appear as dummy-variable contrasts.

vif_diagnostics <- calculate_vif_from_model(
  linear_unstandardized_models[["cesd_full_score"]]
)


################################################################################
# 12. Logistic separation diagnostics
################################################################################

logistic_warning_diagnostics <- map_dfr(
  binary_outcomes,
  function(y) {
    
    warnings_y <- logistic_fit_objects[[y]]$warnings
    
    if (length(warnings_y) == 0) {
      tibble(
        outcome = y,
        measure = measure_label(y),
        warning_message = "No glm warnings"
      )
    } else {
      tibble(
        outcome = y,
        measure = measure_label(y),
        warning_message = warnings_y
      )
    }
  }
)

separation_diagnostics <- map_dfr(
  binary_outcomes,
  function(y) {
    check_categorical_separation(
      data = analytic_sample,
      binary_outcome = y,
      categorical_vars = categorical_predictors
    )
  }
)


################################################################################
# 13. Prepare data for Figure 4A: standardized linear coefficients
################################################################################

figure4a_data <- linear_standardized_results %>%
  filter(term != "(Intercept)") %>%
  mutate(
    measure = factor(
      measure,
      levels = c(
        "Full 19-item modified CES-D",
        "Four-item brief screener",
        "Single depressed-mood item"
      )
    )
  )

term_order_linear <- figure4a_data %>%
  distinct(term, term_label) %>%
  pull(term_label)

figure4a_data <- figure4a_data %>%
  mutate(
    term_label = factor(
      term_label,
      levels = rev(term_order_linear)
    )
  )


################################################################################
# 14. Figure 4A: standardized linear regression coefficients
################################################################################

figure4a_linear <- ggplot(
  figure4a_data,
  aes(
    x = estimate,
    y = term_label,
    xmin = conf.low,
    xmax = conf.high,
    color = measure
  )
) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed"
  ) +
  geom_pointrange(
    position = position_dodge(width = 0.6)
  ) +
  labs(
    title = "Standardized Linear Regression Coefficients Across Measures",
    x = "Standardized coefficient",
    y = NULL,
    color = "Measurement approach",
    caption = "Points are estimates; horizontal lines are 95% Wald confidence intervals."
  ) +
  project_figure_theme +
  theme(
    axis.text.y = element_text(
      size = 10,
      face = "bold",
      colour = "black"
    )
  )

ggsave(
  filename = "outputs/figures/figure4a_linear_standardized_coefficients.png",
  plot = figure4a_linear,
  width = 16,
  height = 8,
  dpi = 300
)


################################################################################
# 15. Prepare data for Figure 4B: logistic odds ratios
################################################################################

figure4b_data <- logistic_results %>%
  filter(term != "(Intercept)") %>%
  mutate(
    measure = factor(
      measure,
      levels = c(
        "Full 19-item modified CES-D",
        "Four-item brief screener",
        "Single depressed-mood item"
      )
    )
  )

term_order_logistic <- figure4b_data %>%
  distinct(term, term_label) %>%
  pull(term_label)

figure4b_data <- figure4b_data %>%
  mutate(
    term_label = factor(
      term_label,
      levels = rev(term_order_logistic)
    )
  )


################################################################################
# 16. Figure 4B: logistic regression odds ratios
################################################################################

figure4b_logistic <- ggplot(
  figure4b_data,
  aes(
    x = odds_ratio,
    y = term_label,
    xmin = conf.low,
    xmax = conf.high,
    color = measure
  )
) +
  geom_vline(
    xintercept = 1,
    linetype = "dashed"
  ) +
  geom_pointrange(
    position = position_dodge(width = 0.6)
  ) +
  scale_x_log10() +
  labs(
    title = "Logistic Regression Odds Ratios Across Measures",
    x = "Odds ratio, log scale",
    y = NULL,
    color = "Measurement approach",
    caption = "Points are odds ratios; horizontal lines are 95% Wald confidence intervals."
  ) +
  project_figure_theme +
  theme(
    axis.text.y = element_text(
      size = 10,
      face = "bold",
      colour = "black"
    )
  )

ggsave(
  filename = "outputs/figures/figure4b_logistic_odds_ratios.png",
  plot = figure4b_logistic,
  width = 16,
  height = 8,
  dpi = 300
)


################################################################################
# 17. Save model objects
################################################################################

regression_model_objects <- list(
  linear_unstandardized_models = linear_unstandardized_models,
  linear_standardized_models = linear_standardized_models,
  logistic_models = logistic_models
)

saveRDS(
  regression_model_objects,
  "data/regression_model_objects.rds"
)


################################################################################
# 18. Save regression outputs
################################################################################

write_xlsx(
  list(
    regression_sample_check = regression_sample_check,
    binary_outcome_check = binary_outcome_check,
    reference_level_table = reference_level_table,
    linear_unstandardized_results = linear_unstandardized_results,
    linear_standardized_results = linear_standardized_results,
    logistic_odds_ratio_results = logistic_results,
    model_fit_summary = model_fit_summary,
    vif_diagnostics = vif_diagnostics,
    logistic_warning_diagnostics = logistic_warning_diagnostics,
    separation_diagnostics = separation_diagnostics
  ),
  "outputs/05_regression_outputs.xlsx"
)


################################################################################
# 19. Print key outputs to console
################################################################################

regression_sample_check
binary_outcome_check
reference_level_table
vif_diagnostics
logistic_warning_diagnostics
separation_diagnostics
model_fit_summary
figure4a_linear
figure4b_logistic
```


# Step 7: Classification agreement and Figure 2

```{r step-06}
################################################################################
# Script: 06_agreement.R
# Purpose: Compute classification agreement across depressive symptom measures
#          and create Figure 2 for the Add Health P1 project
################################################################################




################################################################################
# 1. Read thresholded analytic dataset
################################################################################

analytic_sample <- readRDS("data/addhealth_analytic_sample_thresholded.rds")


################################################################################
# 2. Define binary classification variables
################################################################################

binary_vars <- c(
  "cesd_full_elevated",
  "cesd_brief_elevated",
  "cesd_single_depressed_elevated"
)

score_vars <- c(
  "cesd_full_score",
  "cesd_brief_score",
  "cesd_single_depressed_ord"
)


################################################################################
# 3. Basic quality checks
################################################################################

required_vars <- c(binary_vars, score_vars)

missing_required_vars <- setdiff(required_vars, names(analytic_sample))

if (length(missing_required_vars) > 0) {
  stop(
    paste(
      "These required variables are missing:",
      paste(missing_required_vars, collapse = ", ")
    )
  )
}

agreement_sample_check <- tibble(
  analytic_n = nrow(analytic_sample),
  n_missing_full_binary = sum(is.na(analytic_sample$cesd_full_elevated)),
  n_missing_brief_binary = sum(is.na(analytic_sample$cesd_brief_elevated)),
  n_missing_single_binary = sum(is.na(analytic_sample$cesd_single_depressed_elevated))
)

binary_value_check <- analytic_sample %>%
  summarise(
    across(
      all_of(binary_vars),
      ~ paste(sort(unique(.)), collapse = ", ")
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "binary_variable",
    values_to = "observed_values"
  ) %>%
  mutate(
    expected_values = "0, 1",
    values_correct = observed_values == expected_values
  )


################################################################################
# 4. Define measure pairs
################################################################################

measure_pairs <- tribble(
  ~comparison, ~measure1_label, ~measure1_binary, ~measure2_label, ~measure2_binary,
  "Full vs brief",
  "Full 19-item modified CES-D",
  "cesd_full_elevated",
  "Four-item brief screener",
  "cesd_brief_elevated",
  
  "Full vs single",
  "Full 19-item modified CES-D",
  "cesd_full_elevated",
  "Single depressed-mood item",
  "cesd_single_depressed_elevated",
  
  "Brief vs single",
  "Four-item brief screener",
  "cesd_brief_elevated",
  "Single depressed-mood item",
  "cesd_single_depressed_elevated"
)


################################################################################
# 5. Helper functions
################################################################################

class_label <- function(x) {
  if_else(
    x == 1,
    "Elevated",
    "Not elevated"
  )
}

landis_koch_label <- function(kappa_value) {
  case_when(
    kappa_value <= 0.20 ~ "Poor agreement",
    kappa_value > 0.20 & kappa_value <= 0.40 ~ "Fair agreement",
    kappa_value > 0.40 & kappa_value <= 0.60 ~ "Moderate agreement",
    kappa_value > 0.60 & kappa_value <= 0.80 ~ "Substantial agreement",
    kappa_value > 0.80 ~ "Near-perfect agreement",
    TRUE ~ NA_character_
  )
}

compute_kappa <- function(data, comparison, measure1_label, measure1_binary,
                          measure2_label, measure2_binary) {
  
  x <- data[[measure1_binary]]
  y <- data[[measure2_binary]]
  
  complete_index <- !is.na(x) & !is.na(y)
  
  x <- x[complete_index]
  y <- y[complete_index]
  
  # Create a 2x2 table with fixed category order
  kappa_table <- table(
    factor(x, levels = c(0, 1), labels = c("Not elevated", "Elevated")),
    factor(y, levels = c(0, 1), labels = c("Not elevated", "Elevated"))
  )
  
  # Use psych::cohen.kappa() as specified in the analysis plan
  kappa_object <- psych::cohen.kappa(kappa_table)
  
  # Extract kappa from the psych output
  kappa_value <- as.numeric(kappa_object$kappa)
  
  # Pull the 2x2 counts
  n00 <- as.numeric(kappa_table["Not elevated", "Not elevated"])
  n01 <- as.numeric(kappa_table["Not elevated", "Elevated"])
  n10 <- as.numeric(kappa_table["Elevated", "Not elevated"])
  n11 <- as.numeric(kappa_table["Elevated", "Elevated"])
  
  n_total <- n00 + n01 + n10 + n11
  
  observed_agreement <- (n00 + n11) / n_total
  
  tibble(
    comparison = comparison,
    measure1 = measure1_label,
    measure2 = measure2_label,
    n_total = n_total,
    
    n_agree_not_elevated = n00,
    n_measure2_only = n01,
    n_measure1_only = n10,
    n_agree_elevated = n11,
    
    n_concordant = n00 + n11,
    n_discordant = n01 + n10,
    
    observed_agreement = round(observed_agreement, 4),
    cohen_kappa = round(kappa_value, 4),
    
    landis_koch_interpretation = landis_koch_label(kappa_value),
    
    decision_note = case_when(
      kappa_value > 0.80 ~
        "Measures are largely interchangeable for this pair.",
      kappa_value < 0.40 ~
        "Potential headline finding: low classification agreement.",
      TRUE ~
        "Agreement is not near-perfect; inspect reclassification table."
    )
  )
}

make_reclassification_long <- function(data, comparison, measure1_label,
                                       measure1_binary, measure2_label,
                                       measure2_binary) {
  
  data %>%
    transmute(
      measure1_status = factor(
        class_label(.data[[measure1_binary]]),
        levels = c("Not elevated", "Elevated")
      ),
      measure2_status = factor(
        class_label(.data[[measure2_binary]]),
        levels = c("Not elevated", "Elevated")
      )
    ) %>%
    count(
      measure1_status,
      measure2_status,
      name = "n"
    ) %>%
    complete(
      measure1_status,
      measure2_status,
      fill = list(n = 0)
    ) %>%
    mutate(
      comparison = comparison,
      measure1 = measure1_label,
      measure2 = measure2_label,
      percent_total = round(100 * n / sum(n), 2),
      classification_pattern = case_when(
        measure1_status == "Not elevated" & measure2_status == "Not elevated" ~
          "Concordant: neither measure classifies elevated",
        measure1_status == "Elevated" & measure2_status == "Elevated" ~
          "Concordant: both measures classify elevated",
        measure1_status == "Elevated" & measure2_status == "Not elevated" ~
          paste0("Discordant: ", measure1_label, " only"),
        measure1_status == "Not elevated" & measure2_status == "Elevated" ~
          paste0("Discordant: ", measure2_label, " only"),
        TRUE ~ NA_character_
      )
    ) %>%
    select(
      comparison,
      measure1,
      measure2,
      measure1_status,
      measure2_status,
      classification_pattern,
      n,
      percent_total
    )
}

make_reclassification_matrix <- function(data, measure1_binary,
                                         measure2_binary) {
  
  x <- factor(
    class_label(data[[measure1_binary]]),
    levels = c("Not elevated", "Elevated")
  )
  
  y <- factor(
    class_label(data[[measure2_binary]]),
    levels = c("Not elevated", "Elevated")
  )
  
  tab <- table(x, y)
  
  as.data.frame.matrix(tab) %>%
    rownames_to_column("measure1_status") %>%
    as_tibble()
}


################################################################################
# 6. Cohen's kappa for all pairwise comparisons
################################################################################

kappa_summary <- pmap_dfr(
  measure_pairs,
  function(comparison, measure1_label, measure1_binary,
           measure2_label, measure2_binary) {
    
    compute_kappa(
      data = analytic_sample,
      comparison = comparison,
      measure1_label = measure1_label,
      measure1_binary = measure1_binary,
      measure2_label = measure2_label,
      measure2_binary = measure2_binary
    )
  }
)

kappa_decision_summary <- tibble(
  all_pairs_above_0_80 = all(kappa_summary$cohen_kappa > 0.80),
  any_pair_below_0_40 = any(kappa_summary$cohen_kappa < 0.40),
  overall_decision_note = case_when(
    all_pairs_above_0_80 ~
      "All kappas are above 0.80. Classification results suggest measures are largely interchangeable; emphasize regression and subgroup differences.",
    any_pair_below_0_40 ~
      "At least one kappa is below 0.40. Low classification agreement should be treated as a headline finding.",
    TRUE ~
      "Classification agreement is mixed. Emphasize both prevalence shifts and reclassification patterns."
  )
)


################################################################################
# 7. 2x2 reclassification tables
################################################################################

reclassification_long <- pmap_dfr(
  measure_pairs,
  function(comparison, measure1_label, measure1_binary,
           measure2_label, measure2_binary) {
    
    make_reclassification_long(
      data = analytic_sample,
      comparison = comparison,
      measure1_label = measure1_label,
      measure1_binary = measure1_binary,
      measure2_label = measure2_label,
      measure2_binary = measure2_binary
    )
  }
)

full_vs_brief_2x2 <- make_reclassification_matrix(
  data = analytic_sample,
  measure1_binary = "cesd_full_elevated",
  measure2_binary = "cesd_brief_elevated"
) %>%
  rename(
    `Full scale status` = measure1_status
  )

full_vs_single_2x2 <- make_reclassification_matrix(
  data = analytic_sample,
  measure1_binary = "cesd_full_elevated",
  measure2_binary = "cesd_single_depressed_elevated"
) %>%
  rename(
    `Full scale status` = measure1_status
  )

brief_vs_single_2x2 <- make_reclassification_matrix(
  data = analytic_sample,
  measure1_binary = "cesd_brief_elevated",
  measure2_binary = "cesd_single_depressed_elevated"
) %>%
  rename(
    `Brief screener status` = measure1_status
  )


################################################################################
# 8. Prevalence table for Figure 2 Panel 1
################################################################################

prevalence_summary <- tibble(
  measure = c(
    "Full 19-item modified CES-D",
    "Four-item brief screener",
    "Single depressed-mood item"
  ),
  binary_variable = c(
    "cesd_full_elevated",
    "cesd_brief_elevated",
    "cesd_single_depressed_elevated"
  ),
  n_total = nrow(analytic_sample),
  n_elevated = c(
    sum(analytic_sample$cesd_full_elevated == 1),
    sum(analytic_sample$cesd_brief_elevated == 1),
    sum(analytic_sample$cesd_single_depressed_elevated == 1)
  )
) %>%
  mutate(
    percent_elevated = round(100 * n_elevated / n_total, 2)
  )


################################################################################
# 9. Classification overlap table for Figure 2 Panel 2
################################################################################

classification_overlap_summary <- analytic_sample %>%
  mutate(
    n_elevated_measures =
      cesd_full_elevated +
      cesd_brief_elevated +
      cesd_single_depressed_elevated,
    
    classification_overlap = case_when(
      n_elevated_measures == 0 ~ "Not elevated on any measure",
      n_elevated_measures == 1 ~ "Elevated on one measure",
      n_elevated_measures == 2 ~ "Elevated on two measures",
      n_elevated_measures == 3 ~ "Elevated on all three measures",
      TRUE ~ NA_character_
    ),
    
    classification_overlap = factor(
      classification_overlap,
      levels = c(
        "Not elevated on any measure",
        "Elevated on one measure",
        "Elevated on two measures",
        "Elevated on all three measures"
      )
    )
  ) %>%
  count(
    classification_overlap,
    name = "n"
  ) %>%
  mutate(
    n_total = sum(n),
    percent = round(100 * n / n_total, 2)
  )


################################################################################
# 10. Figure 2: Prevalence and classification overlap
################################################################################

figure2_panel1_data <- prevalence_summary %>%
  transmute(
    panel = "Panel A. Prevalence by measure",
    category = measure,
    n = n_elevated,
    percent = percent_elevated,
    label = paste0(round(percent, 1), "%")
  )

figure2_panel2_data <- classification_overlap_summary %>%
  transmute(
    panel = "Panel B. Classification overlap",
    category = as.character(classification_overlap),
    n = n,
    percent = percent,
    label = paste0(round(percent, 1), "%")
  )

figure2_data <- bind_rows(
  figure2_panel1_data,
  figure2_panel2_data
) %>%
  mutate(
    panel = factor(
      panel,
      levels = c(
        "Panel A. Prevalence by measure",
        "Panel B. Classification overlap"
      )
    )
  )

figure2_prevalence_overlap <- ggplot(
  figure2_data,
  aes(
    x = category,
    y = percent,
    fill = category
  )
) +
  geom_col(
    width = 0.6,
    show.legend = FALSE
  ) +
geom_text(
  aes(label = label),
  vjust = -0.25,
  size = 4,
  fontface = "bold",
  colour = "black"
) +
  facet_grid(
    . ~ panel,
    scales = "free_x",
    space = "free_x"
  ) +
  scale_x_discrete(
    labels = function(x) stringr::str_wrap(x, width = 16)
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.12))
  ) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Prevalence and Classification Differences Across Measures",
    x = NULL,
    y = "Percent of respondents") +
project_figure_theme

ggsave(
  filename = "outputs/figures/figure2_prevalence_classification_overlap.png",
  plot = figure2_prevalence_overlap,
  width = 13,
  height = 6,
  dpi = 300
)


################################################################################
# 11. Save agreement outputs
################################################################################

write_xlsx(
  list(
    agreement_sample_check = agreement_sample_check,
    binary_value_check = binary_value_check,
    kappa_summary = kappa_summary,
    kappa_decision_summary = kappa_decision_summary,
    reclassification_long = reclassification_long,
    full_vs_brief_2x2 = full_vs_brief_2x2,
    full_vs_single_2x2 = full_vs_single_2x2,
    brief_vs_single_2x2 = brief_vs_single_2x2,
    prevalence_summary = prevalence_summary,
    classification_overlap_summary = classification_overlap_summary
  ),
  "outputs/06_agreement_outputs.xlsx"
)


################################################################################
# 12. Print key outputs to console
################################################################################

agreement_sample_check
binary_value_check
kappa_summary
kappa_decision_summary
reclassification_long
prevalence_summary
classification_overlap_summary
figure2_prevalence_overlap
```

The following code generates the descriptive Table 1 presented in the
manuscript. It is included here to maintain a complete reproducible record of
all manuscript tables.


# Step 8: Manuscript Table 1



```{r manuscript-table-1}
################################################################################
# Manuscript Table 1: Characteristics of the analytic sample
################################################################################

library(gtsummary)
library(gt)
library(dplyr)

table1 <- readRDS(
  "data/addhealth_analytic_sample_thresholded.rds"
) %>%
  select(
    age,
    gender,
    race_eth,
    resident_parent_education,
    family_structure,
    school_connectedness,
    peer_support,
    cesd_full_score,
    cesd_brief_score,
    cesd_single_depressed_ord
  ) %>%
  tbl_summary(
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(
      all_continuous() ~ 2,
      all_categorical() ~ c(0, 1)
    ),
    label = list(
      age ~ "Age, years",
      gender ~ "Gender",
      race_eth ~ "Race/ethnicity",
      resident_parent_education ~
        "Highest resident-parent education",
      family_structure ~ "Family structure",
      school_connectedness ~
        "School connectedness score",
      peer_support ~
        "Peer social support score",
      cesd_full_score ~
        "Full 19-item modified CES-D score",
      cesd_brief_score ~
        "Four-item affective screener score",
      cesd_single_depressed_ord ~
        "Single depressed-mood item"
    ),
    missing = "no"
  ) %>%
  modify_header(
    label = "**Characteristic**",
    all_stat_cols() ~ "**Overall, N = {N}**"
  ) %>%
  bold_labels()

table1 %>%
  as_gt() %>%
  gt::tab_source_note(
    gt::md(
      paste(
        "*Note.* Values are mean (SD) for continuous variables and",
        "n (%) for categorical variables. Higher school-connectedness",
        "scores indicate weaker connectedness.",
        "CES-D = Center for Epidemiologic Studies Depression Scale."
      )
    )
  ) %>%
  gt::tab_options(
    table.font.size = gt::px(9),
    data_row.padding = gt::px(2),
    table.width = gt::pct(100)
  )
```
```

